---
title: "UMAP with sample density"
output: html_notebook
---

# Goal
Take a seurat object and make a custom UMAP panels showing the density of each sample throughout the UMAP embedding.

# Dependencies
```{r, results='hide'}
library(Seurat)
library(tidyverse)
library(viridisLite)
```

# Load the data
We'll look at a scRNA-seq time course of the prostate cancer cell line DU145 treated with TGFB1 for different lengths of time. This data is originally from [this paper](https://www.nature.com/articles/s41467-020-16066-2), but we'll download it from Figshare:

```{r}
seu <- readRDS(gzcon(url("https://ndownloader.figshare.com/files/27965040")))
```

```{r}
DimPlot(seu, group.by="Sample", reduction='umap_pseudo')
```

# Density plot
We'll get the relevant data into a data frame. We just need the meta.data table and UMAP embeddings
```{r}
dat <- data.frame(seu@meta.data)
dat$UMAP1 <- Embeddings(seu, "umap_pseudo")[,1]
dat$UMAP2 <- Embeddings(seu, "umap_pseudo")[,2]
```

Our plot will consistent of two elements: individual points that will all be the same color, and the layer containing each sample's density on the plot.

```{r}
#Create a data frame that doesn't contain a "sample" column. This will allow us to facet the density layer without affecting the points
dat_bg <- dat[,-(which(colnames(dat)=="Sample"))]
```

And the sample names are long right now, so I'll just trim them:
```{r}
dat$Sample <- gsub("DU145_TGFB1_", "", dat$Sample)
dat$Sample <- factor(dat$Sample, levels=c("0d", "8h", "1d", "3d", "7d", "8h_rm", "1d_rm", "3d_rm"))
```


Now we're ready to plot

```{r}
density_plot <- ggplot(dat, aes(x=UMAP1, y=UMAP2)) +
  stat_density_2d(geom="raster", aes(fill=stat(ndensity)), contour=F) + #ndensity calculates the normalized density for each sample--otherwise density would be affected by the number of cells for each sample, which is variable
  geom_point(data=dat_bg, shape=16, size=0.1, alpha=0.2, color="white") +
  scale_fill_gradientn(colours=viridisLite::mako(100), name="Density") +
  scale_x_continuous(expand=c(0,0)) +
  scale_y_continuous(expand=c(0,0)) +
  facet_wrap(~Sample, ncol=8) +
  theme_classic() +
  theme(strip.background = element_blank(),
        strip.text = element_text(size=12, color="black"),
        axis.text=element_blank(),
        axis.title=element_blank(),
        axis.ticks=element_blank(),
        axis.line = element_blank(),
        plot.background = element_rect(fill = "transparent", color = NA),
        legend.text=element_text(size=12, color="black"),
        legend.title=element_text(size=14, color="black"))
```

```{r, fig.width=10, fig.height=1.8}
density_plot
```

Save image
```{r}
ggsave(density_plot,
       filename="umap_sample_density.png",
       dpi=150,
       width=10, height=1.8)
```

# Session Info
```{r}
sessionInfo()
```


