---
title: "Seurat's FeaturePlot from scratch"
output: github_document
---

# Goal
Seurat is an amazing tool to handle scRNA-seq data. When preparing plots for publication, it can be nice to have a bit more control over the plot. I'm a big fan of ggplot2, which is the plotting tool Seurat uses under the hood. Here I'll use it to rebuild the standard FeaturePlot().

# Dependencies
```{r}
library(Seurat)
library(SeuratData)
library(tidyverse) #loads ggplot2
library(ggrastr)
options(ggrastr.default.dpi=300)
```

# Load data
For convenience, we'll use some datasets included in the [SeuratData](https://github.com/satijalab/seurat-data) package. 
```{r}
InstallData("pbmc3k")
data("pbmc3k")
seu <- pbmc3k
```

## Quick processing
```{r, results=FALSE, message=FALSE, warning=FALSE}
seu <- NormalizeData(seu) #just to store log-transformed counts
seu <- SCTransform(seu)
seu <- RunPCA(seu, verbose=F)
seu <- RunUMAP(seu, dims=1:30)
seu <- FindNeighbors(seu, dims=1:30)
seu <- FindClusters(seu, resolution=0.2)
```

```{r}
DimPlot(seu, label=T)
```

# Seurat's FeaturePlot
To its credit, Seurat's FeaturePlot does have a reasonable amount of customizability (see the function's help menu with ?FeaturePlot). However, I find I hit walls when trying to get the aesthetics just right for presentations/publications

```{r}
DefaultAssay(seu) <- "RNA" #we'll visualize the log-transformed counts
```


```{r}
#Default parameters
FeaturePlot(seu, features="CD3E")
```

So this is totally fine. But if I was getting picky, there are a few things I would change (in no particular order):
- Remove underscores in the axis titles
- Maybe remove the axes all together for a cleaner visualization
- I'm not a fan of the grey-to-blue color map
- I'm not crazy about the white ticks ggplot2 uses in the legend by default
- I typically don't like bold text the plots (the title), and I've seen at least one journal not allow it
- Add legend title (what are we quantifying??)
- Would also be nice to have freedom to modify point size/transparency, and relative font sizes.

# Custom FeaturePlot with ggplot2
Ggplot2 takes a data frame as input. For this plot, the only variables we need are the embedding coordinates and the expression values of a given gene. Note that this could be extended to mimick the categorical DimPlot function by replacing expression values with some categorical data (eg. cluster)

Ggplot2 code can be lengthy, so I'll write the plotting code up as a function that takes a gene symbol as input. That way we can just re-use the code for as many genes as we want.

```{r}
plotGene <- function(gene){
  exp_dat <- data.frame(UMAP1 = Embeddings(seu, "umap")[,1],
                        UMAP2 = Embeddings(seu, "umap")[,2],
                        Expression = seu[["RNA"]]@data[gene,]) #pulls log-transformed scaled counts out
  
  # The commented out code in the next line orders the data frame from lowest to highest expression
  # As a result, the highest expressing cells get added to the "top" of the plot
  # This is common and is sometimes helpful, but given how over-plotted many UMAP plots are,
  # this could give false impressions about the distribution of the gene's expression.
  # Remove hash below if you want, but use with caution:
  #exp_dat <- exp_dat[order(exp_dat$Expression),]
  
  exp_plot <- ggplot(exp_dat, aes(x=UMAP1, y=UMAP2)) +
    geom_point_rast( #this comes from ggrastr--rasterizes the points to avoid large vector graphics
      size=0.25, #make smaller when working with more cells
      shape=16, #uses a specific point shape that doesn't have an outline
      alpha=0.75, # a little transparency helps with overplotting
      aes(color=Expression)) +
    scale_color_gradientn(colours=c("lightgrey", rev(viridis::inferno(100))), #inferno for non-zero values
                          name="log(TP10k)",
                          guide = guide_colorbar(ticks.colour = "black",
                                              frame.colour = "black",
                                               barwidth=0.75)) +
    ggtitle(gene) +
    theme_classic() +
    theme(axis.text = element_text(size=10, color="black"),
          axis.title = element_text(size=12),
          legend.text = element_text(size=10),
          legend.title = element_text(size=10),
          plot.title = element_text(size=14, face = "italic"))
  return(exp_plot)
}
```

Let's try it out

```{r}
plotGene("CD3E")
```

Looks good! All of those plot elements can be modified to fit your taste.

## Save plot
```{r}
p1 <- plotGene("CD3E")
ggsave(p1, filename="./umap_expression_cd3e.pdf",
       width=3.25, height=2.5)
```

# Session Info
```{r}
sessionInfo()
```

